---
title: "Analysis of Housing train"
author: "Tilon Bobb"
date: "May 08, 2024"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## train Loading and Initial Exploration

We begin by loading the trainset from a URL and taking a preliminary
look at its structure.

```{r load-train}
train <- read.csv("https://raw.githubusercontent.com/Kingtilon1/House-prices-competition/main/train.csv", stringsAsFactors = FALSE)
head(train)
summary(train)
```

## Identification of Variables

We identify quantitative variables and check for skewness, choosing
`LotArea` as $X$ and `SalePrice` as $Y$.

```{r identify-variables}
library(ggplot2)
ggplot(train, aes(x=LotArea)) + geom_histogram(bins=30, fill="blue", color="black") +
  ggtitle("Distribution of LotArea") + theme_minimal()
ggplot(train, aes(x=SalePrice)) + geom_histogram(bins=30, fill="red", color="black") +
  ggtitle("Distribution of SalePrice") + theme_minimal()
```

## Calculation of Probabilities and Independence Check

### Calculation of Quartiles and Probabilities

We calculate the necessary quartiles and probabilities.

```{r calculate-probabilities}
x_third_quartile <- quantile(train$LotArea, 0.75)
y_second_quartile <- quantile(train$SalePrice, 0.5)

condition_X_greater_x <- train$LotArea > x_third_quartile
condition_Y_greater_y <- train$SalePrice > y_second_quartile

prob_X_greater_x <- mean(condition_X_greater_x)
prob_Y_greater_y <- mean(condition_Y_greater_y)
prob_X_greater_x_given_Y_greater_y <- mean(train$LotArea[condition_Y_greater_y] > x_third_quartile)
prob_X_greater_x_and_Y_greater_y <- mean(condition_X_greater_x & condition_Y_greater_y)
prob_X_less_x_given_Y_greater_y <- mean(train$LotArea[condition_Y_greater_y] < x_third_quartile)

cat("P(X > x): ", prob_X_greater_x, "\nP(Y > y): ", prob_Y_greater_y, "\nP(X > x | Y > y): ", prob_X_greater_x_given_Y_greater_y,
    "\nP(X > x and Y > y): ", prob_X_greater_x_and_Y_greater_y, "\nP(X < x | Y > y): ", prob_X_less_x_given_Y_greater_y)
```

1: P(X\>x)=0.25: This probability indicates that 25% of the properties
have a lot area greater than the third quartile of all lot areas in the
dataset. This is by definition of the quartile, as the third quartile
(75th percentile) is the value below which 75% of the data fall. 𝑃 ( 𝑌
\> 𝑦 ) = 0.4986301

2: P(Y\>y)=0.4986301: About 49.9% of the properties have a sale price
higher than the median (second quartile, 50th percentile) sale price.
This is very close to 50%, as expected, because the median is the middle
value in a data set.

3: 𝑃 ( 𝑋 \> 𝑥 ∣ 𝑌 \> 𝑦 ) = 0.3791209 P(X\>x∣Y\>y)=0.3791209: Given that
a property's sale price is above the median, there is approximately a
37.9% chance that its lot area is also above the third quartile. This
suggests that among higher-priced homes, a significantly large
proportion also have larger lot areas, though not the majority.

4: 𝑃 ( 𝑋 \> 𝑥 ∧ 𝑌 \> 𝑦 ) = 0.1890411 P(X\>x∧Y\>y)=0.1890411: There is an
18.9% chance that a property will have both a sale price above the
median and a lot area greater than the third quartile. This joint
probability is less than the product of the individual probabilities ( 𝑃
( 𝑋 \> 𝑥 ) × 𝑃 ( 𝑌 \> 𝑦 ) P(X\>x)×P(Y\>y)), indicating a possible
dependency between lot area and sale price --- large lot areas tend to
occur with higher prices, but not as frequently as might be expected if
the two were independent.

5: 𝑃 ( 𝑋 \< 𝑥 ∣ 𝑌 \> 𝑦 ) = 0.6208791 P(X\<x∣Y\>y)=0.6208791: Given a
property's sale price is above the median, there is a 62.1% chance that
its lot area is less than the third quartile. This indicates that even
among higher-priced homes, it's more common for the lot area to be in
the smaller three-quarters of all lot areas. \### Chi-Square Test for
Independence

# Creating a table of counts

## Additional Analysis: Contingency Table of Counts

To further analyze the relationship between `LotArea` (X) and
`SalePrice` (Y), we create a contingency table that categorizes
properties based on whether they fall above or below these quartiles.

```{r contingency-table}
train$X_group <- ifelse(train$LotArea > x_third_quartile, ">3rd quartile", "<=3rd quartile")
train$Y_group <- ifelse(train$SalePrice > y_second_quartile, ">2nd quartile", "<=2nd quartile")

count_table <- table(train$X_group, train$Y_group)
addmargins(count_table)
```

## Interpretation of Contingency Table

The contingency table reveals insightful patterns about the distribution
of `LotArea` and `SalePrice` among the properties:

-   **Properties with `LotArea` \<= 3rd Quartile and `SalePrice` \<= 2nd
    Quartile (643 properties)**: This is the most populous category,
    indicating that the majority of properties in the dataset feature
    lot areas that are smaller or up to the median size and are priced
    at or below the median sale price. These properties represent a
    typical, more affordable segment of the housing market.

-   **Properties with `LotArea` \<= 3rd Quartile and `SalePrice` \> 2nd
    Quartile (452 properties)**: A significant number of properties have
    smaller or average-sized lots but are priced above the median. This
    suggests that factors other than lot size, such as location, home
    features, or market conditions, may be contributing to higher
    property values in this group.

-   **Properties with `LotArea` \> 3rd Quartile and `SalePrice` \<= 2nd
    Quartile (89 properties)**: Fewer properties fall into this category
    where larger lot sizes do not correspond to higher prices, possibly
    indicating underdeveloped areas or locales where land is less of a
    premium factor in determining house prices.

-   **Properties with `LotArea` \> 3rd Quartile and `SalePrice` \> 2nd
    Quartile (276 properties)**: These properties represent a premium
    segment of the market, where both larger lot sizes and higher prices
    coincide. This could suggest more desirable locations or luxury
    estates where buyers are willing to pay a premium for more space.

The distribution highlights the nuanced relationship between lot size
and property price, showing that while there is a tendency for larger
lots to fetch higher prices, many properties defy this trend due to
other influencing factors.

```{r chi-square-test}
table_A_B <- table(condition_X_greater_x, condition_Y_greater_y)
chisq.test(table_A_B)
```

### Analysis of Independence Using Chi-Square Test

The Chi-Square test results provide a strong indication of the
relationship between `LotArea` (X) and `SalePrice` (Y) when split by the
3rd and 2nd quartiles, respectively. The test yields a Chi-Square
statistic of 127.74 with a p-value significantly less than 0.05 (p-value
\< 2.2e-16), which strongly rejects the null hypothesis of independence.
This means that splitting the data by these quartiles does not result in
independent subsets; rather, there is a significant association between
larger lot areas and higher sale prices.

Additionally, when we compare the calculated probabilities: - $P(A|B)$
(the probability of $X > x$ given $Y > y$) is 0.3791209. - $P(A)P(B)$
(the product of the probabilities $P(X > x)$ and $P(Y > y)$), which is
calculated as $0.25 \times 0.4986301 = 0.1246575$.

The inequality $P(A|B) \neq P(A)P(B)$ indicates a dependency between the
variables, where properties with a higher sale price are more likely to
also have a larger lot area than would be expected under the condition
of independence. The result of this mathematical check aligns with the
Chi-Square test, further affirming that the variables are not
independent.

In summary, the statistical evidence from both the probability
calculations and the Chi-Square test confirms that the manner in which
the training data has been split (based on quartiles of `LotArea` and
`SalePrice`) leads to a dependent relationship between the two
variables. This dependency should be considered when analyzing or
modeling these real estate data.

### Univariate Descriptive Statistics and Plots

We start by providing basic descriptive statistics for the entire
dataset, focusing particularly on `LotArea` and `SalePrice`.

```{r descriptive-stats}
summary(train$LotArea)
summary(train$SalePrice)
```

The summaries of LotArea and SalePrice illustrate key characteristics of
the real estate market in the dataset. The LotArea ranges significantly
from 1,300 to 215,245 square feet, indicating diverse property sizes,
while the SalePrice varies from \$34,900 to \$755,000, reflecting a wide
economic spread in property values. Both distributions are right-skewed,
as indicated by means higher than medians, typical for real estate where
a few high values can skew the average upward. These statistics are
vital for understanding property size and pricing dynamics within the
market.

```{r scatterplot}
plot(train$LotArea, train$SalePrice, main="Scatterplot of LotArea vs SalePrice",
     xlab="LotArea", ylab="SalePrice", pch=19, col=rgb(0.1, 0.2, 0.5, 0.7))
```

The scatter plot of LotArea versus SalePrice reveals a broadly positive
relationship, indicating that properties with larger lot areas generally
tend to have higher sale prices. However, the relationship isn't
strictly linear, and there is significant variability, especially among
properties with larger lot areas. Most data points cluster toward the
lower end of both axes, suggesting that smaller, more affordable
properties are more prevalent. Outliers and the spread of data points at
higher lot areas underscore that factors other than lot size, such as
location and property features, also significantly influence sale
prices.

### 3. 95% Confidence Interval for the Difference in the Mean of the Variables

### 95% Confidence Interval for the Difference in Mean

We compute the 95% confidence interval for the difference in the mean of
`LotArea` and `SalePrice`.

```{r ci-difference-mean}
lot_mean <- mean(train$LotArea)
sale_mean <- mean(train$SalePrice)
se_diff <- sqrt(var(train$LotArea)/length(train$LotArea) + var(train$SalePrice)/length(train$SalePrice))
ci_lower <- (lot_mean - sale_mean) - qt(0.975, df=min(length(train$LotArea), length(train$SalePrice))-1) * se_diff
ci_upper <- (lot_mean - sale_mean) + qt(0.975, df=min(length(train$LotArea), length(train$SalePrice))-1) * se_diff
c(ci_lower, ci_upper)
```

The output displays a 95% confidence interval for the difference in
means between LotArea and SalePrice, ranging from approximately
-174,514.8 to -166,293.9. This interval suggests that SalePrice is, on
average, significantly higher than LotArea by this range of values. The
negative sign indicates the direction of the difference due to the
subtraction order in the calculation (LotArea minus SalePrice). This
statistically significant difference highlights distinct scales and
units between the two variables, reaffirming their disparate magnitudes
in the dataset.

### Correlation Matrix and Hypothesis Testing

First,I'll get the correlation matrix for `LotArea` and `SalePrice`.

```{r correlation-matrix}
cor_matrix <- cor(train[,c("LotArea", "SalePrice")])
cor_matrix
```

The output presents a correlation matrix between LotArea and SalePrice,
showing a correlation coefficient of approximately 0.264 between these
two variables. This value indicates a positive but weak correlation,
suggesting that while there is some degree of association where larger
lot areas tend to correlate with higher sale prices, the relationship is
not strongly linear. The coefficients on the diagonal (1.0000000)
confirm that each variable perfectly correlates with itself, as
expected.

Next, lets test the hypothesis that the correlation between LotArea and
SalePrice is 0, using a t-test and provide a 99% confidence interval.

```{r}
cor_test <- cor.test(train$LotArea, train$SalePrice)
cor_test

t_value <- cor_test$estimate / cor_test$std.error
df <- cor_test$parameter
cor_ci_lower <- cor_test$estimate - qt(0.995, df) * cor_test$std.error
cor_ci_upper <- cor_test$estimate + qt(0.995, df) * cor_test$std.error
c(cor_ci_lower, cor_ci_upper)
```

The correlation test performed between LotArea and SalePrice
demonstrates a statistically significant but modest positive correlation
of approximately 0.264, as confirmed by the Pearson's product-moment
correlation test. This result, with a t-value of 10.445 and 1458 degrees
of freedom, leads to a p-value less than 2.2e-16, strongly rejecting the
null hypothesis that no correlation exists between the two variables.
The 99% confidence interval for this correlation, extending from 0.215
to 0.311, solidifies the finding that larger lot areas are generally
associated with higher sale prices, suggesting that while lot size does
impact sale price, other factors also play significant roles in
determining property values. This correlation is indicative of a
relationship where properties with greater lot areas tend to command
higher prices, although the relationship is not overwhelmingly strong.

### Inverting the Correlation Matrix and Matrix Operations

Calculate and invert the correlation matrix, then perform matrix
multiplications:

```{r matrix-operations}
cor_matrix <- cor(train[,c("LotArea", "SalePrice")])
precision_matrix <- solve(cor_matrix)

mult_cor_prec <- cor_matrix %*% precision_matrix

mult_prec_cor <- precision_matrix %*% cor_matrix

mult_cor_prec
```

### Conducting Principal Components Analysis (PCA)

Principal Components Analysis (PCA) is a statistical technique used to
reduce the dimensionality of a dataset, increasing interpretability
while minimizing information loss. It transforms the data into a new
coordinate system, such that the greatest variance comes to lie on the
first few principal axes.

### Principal Components Analysis (PCA)

Conduct PCA on the selected variables and interpret the results:

```{r pca-analysis}
library(stats)
pca_result <- prcomp(train[,c("LotArea", "SalePrice")], scale. = TRUE)

summary(pca_result)

plot(pca_result, type = "lines")
```

The PCA analysis performed on the "LotArea" and "SalePrice" variables
reveals that the first principal component (PC1) accounts for the
majority of the variability in the data (63.19%), with a standard
deviation of approximately 1.1242. PC2 captures additional variability
(36.81%) orthogonal to PC1. Together, PC1 and PC2 explain 100% of the
total variability in the data. This suggests that PC1 represents the
primary trend in the dataset, while PC2 captures secondary patterns not
explained by PC1.

### Preparing Data and Fitting Exponential Distribution

```{r fit-exponential}
library(MASS)

shifted_lot_area <- train$LotArea - min(train$LotArea) + 1

fit <- fitdistr(shifted_lot_area, densfun = "exponential")

lambda <- fit$estimate
lambda
```

### Sample from the Fitted Distribution and Plot Histograms

After fitting the distribution, sample from it and compare the results
with the original data through histograms.

\### Sampling and Plotting

```{r sample-plot}
samples <- rexp(1000, rate = lambda)

hist(samples, main="Histogram of Exponential Samples", col="blue", breaks=30)

hist(shifted_lot_area, main="Histogram of Shifted LotArea", col="red", breaks=30)
```

### Calculate Percentiles and Confidence Intervals

Find percentiles using the exponential probability density function and
generate confidence intervals for empirical data.

### Percentiles and Confidence Intervals

```{r percentiles-ci}
exp_5th <- qexp(0.05, rate = lambda)
exp_95th <- qexp(0.95, rate = lambda)

empirical_5th <- quantile(shifted_lot_area, 0.05)
empirical_95th <- quantile(shifted_lot_area, 0.95)

mean_lot <- mean(train$LotArea)
sd_lot <- sd(train$LotArea)
n_lot <- length(train$LotArea)
ci_lower <- mean_lot - qt(0.975, df=n_lot-1) * sd_lot / sqrt(n_lot)
ci_upper <- mean_lot + qt(0.975, df=n_lot-1) * sd_lot / sqrt(n_lot)

list(exp_5th = exp_5th, exp_95th = exp_95th, empirical_5th = empirical_5th, empirical_95th = empirical_95th, ci_lower = ci_lower, ci_upper = ci_upper)
```

I adjusted LotArea for fitting an exponential distribution, revealing a
λ (lambda) parameter of approximately 0.0001085, indicating a relatively
slow rate of decay. The comparison between histograms of the sampled
data and the actual data shows that while the exponential model roughly
captures the distribution's right skewness, it fails to accurately
represent the distribution's tail behavior, particularly at higher
values. Descriptive statistics highlight this by showing significant
variance in LotArea values, ranging from 1,300 to 215,245 square feet.
The empirical 5th (2,012.7 sq ft) and 95th (16,102.15 sq ft) percentiles
of LotArea contrast with those derived from the exponential model
(472.81 for 5th and 27,614.15 for 95th), underscoring the model's
limitations in predicting extreme values. Moreover, the correlation
between LotArea and SalePrice is moderately weak at 0.264, suggesting
limited linear predictability between lot size and sale price. The 95%
confidence interval for the mean LotArea (10,004.42 to 11,029.24)
further quantifies uncertainty in estimating the average lot size,
reinforcing the need for a more nuanced model to fully capture LotArea
characteristics in the real estate market context. This comprehensive
analysis not only quantifies various statistical properties of LotArea
but also highlights the necessity of selecting appropriate models to
reflect its distribution accurately.

### Building the Regression Model

```{r regression-model}
library(stats)

model <- lm(SalePrice ~ LotArea, data=train)

summary(model)
```
The linear regression model, using LotArea to predict SalePrice, shows that each additional square foot of lot area increases the sale price by approximately $2.10, statistically significant with a p-value less than 2e-16. The model has a relatively low R-squared value of 0.06961, indicating that LotArea alone explains about 6.96% of the variance in SalePrice, suggesting other factors also play significant roles in determining property prices. The residuals indicate that while the model captures central tendencies, there is considerable variability in predictions, with errors ranging from about -$275,668 to $553,356. The intercept value of approximately $158,800 suggests that the base price for the smallest properties in the dataset is substantial. Overall, the model highlights the positive, yet limited influence of LotArea on SalePrice and underscores the need for more complex models to better capture the dynamics of real estate pricing.


```{r}
test <- read.csv('https://raw.githubusercontent.com/Kingtilon1/House-prices-competition/main/test.csv')

predictions <- predict(model, newdata = test)
submission <- data.frame(Id = test$Id, SalePrice = predictions)
write.csv(submission, "submission.csv", row.names = FALSE)
```

